Copulas reveal complex and informative dependencies propagating throughout ecology

### 2019-05-16 Paper.Rmd
minor edits from one final read before submitting to Ecol Lett

\noindent \emph{The number of words in the main text (excluding abstract, acknowledgements, references, table and figure legends):} 7493

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=3.5in]{./Results/PedagogFig1.pdf}
\caption{(A-D) Bivariate datasets showing diverse relationships between 
variables; many of these datasets nevertheless have the same Pearson (P)
or Spearman (S) correlation coefficients. (C, D) Contours 
of probability density functions (pdf) of Clayton (C) and survival-Clayton 
(D) copulas with parameter 5 (see section on methods for the observational hypothesis), together
with samples from the corresponding distributions. (E,F) Reflections
of C, D, respectively, about a vertical axis. (G) Contours of the
pdf function of a normal copula (methods for the observational hypothesis) with 
parameter chosen to
have the same Spearman correlation as A-D, together with a sample.
(H) The portion of the Spearman correlation within a band (diagonal lines
show two bands, the data in the left/lower band showing stronger
dependence than the right/upper band) can be computed 
for any band (methods for the observational hypothesis) to describe how dependence
varies from one part of the two distributions to another.
Subscripts h and v 
stand for "horizontal" and "vertical". \label{fig_cop_pedag1}}
\end{center}
\end{figure}

Well-developed approaches do exist, however, and are 
applied widely in other fields, 
that provide a description of the dependence 
between two or more variables that can be mathematically proven 
to be complete [@joe2014_dependence; @nelsen2006_copula]. 
The approaches are based on 
the concept of the *copula*, which we here 
introduce. Copula approaches separate the information of a bivariate 
random variable into two distinct parts: the information in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
the dependence). From a data perspective, given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), the dependence
structure is revealed by the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The rank
of the smallest element in a set is $1$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$. See, 
for instance,
Fig. \ref{fig_cop_pedag1}C-D (the points), which show the copula structures of 
A-B, respectively. Ranking makes the marginal distributions of 
the underlying random variables uniform, isolating the 
dependence structure.
See @nelsen2006_copula and @joe2014_dependence
for general background. A theorem of @sklar1959_theorem
states that copulas give a unique and complete 
representation of the dependence
between variables.

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=7in]{./MasterFigure.pdf}
\caption{Guide for hypotheses addressed in the text and how each instantiates the
overall hypothesis. Labels are defined 
in the text to refer to parts of the figure. "Causal hypotheses" (i.e., hypotheses about 
causes of copula structure) are labeled with letters A-D. "Consequential 
hypothesis" (i.e., hypothesis about 
consequences of copula structure) are labeled with X-Z. The "observational 
hypothesis" (i.e., our hypothesis about 
copula structure of data) is labeled O.\label{MasterFigure}}
\end{center}
\end{figure}

One way we may expect a copula approach 
to better reflect ecological relationships relates to Liebig's law of the
minimum. If, for instance, the growth of a plant depends on soil 
nitrogen, N, and other factors, a plot of plant growth rates versus N may 
look more like Fig. \ref{fig_cop_pedag1}A or E than like 
Fig. \ref{fig_cop_pedag1}B or F. Nitrogen 
will control plant growth, producing a clear
relationship, only when it is limiting, in the left portions of the distributions. 
The portions of the distributions that dominate the association is visible in the copula
approach, but not via standard techniques. We therefore frame "causal hypothesis" A 
(labeled in Fig. \ref{MasterFigure}): 
copulas of dependencies between environmental and 
ecological variables will sensitively reveal 
nonlinear, asymmetric and threshold effects for population, community, 
and ecosystem functioning variables; copulas between 
environmental and ecological variables
will commonly show tail dependence; such tail dependence will lead to 
tail dependence across space in ecological variables.

@@

Our causal and consequential hypotheses imply, or rely on, an "observational
hypothesis" (O in Fig. \ref{MasterFigure}), which we here frame: datasets in 
ecology will commonly have copula structure
distinct from that of a Gaussian/normal distribution (here called 
non-normal copula structure); they will commonly show tail dependence 
distinct from that of a normal distribution; in particular they will
show asymmetric tail dependence.

@@

Conceptually, a bivariate copula is a random variable 
on the unit square, $[0,1] \times [0,1]$, with uniform maginals. 
If $(X,Y)$ is a continuous random
variable and the cummulative distribution functions of $X$ and $Y$ are 
$F_X(x)=P[X\leq x]$ and $F_Y(y)=P[Y\leq y]$, then the dependence structure
of $(X,Y)$ is revealed by the random variable $(F_X(X),F_Y(Y))$,
which takes values in $[0,1] \times [0,1]$ and has 
uniform marginals. 
We introduce the copula families we used, and 
measures of tail dependence.
The normal copula family (henceforth N) is a one-parameter
family, with parameter corresponding to the correlation of the related 
bivariate normal distribtion, and controlling the degree of dependence 
(Fig. \ref{fig_cop_pedag1}G for an example).
Left- and right-tail dependence are $0$ for the N copula, and this 
serves as a benchmark for tail dependence in other copulas - positive values of our
tail dependence measures correspond to dependence stronger than an N
copula. The Clayton and 
survival-Clayton families (Fig. \ref{fig_cop_pedag1}C-D) are two more one-parameter 
families. The Clayton has positive lower-tail dependence ($0.87$ for
the example of Fig. \ref{fig_cop_pedag1}C) and $0$ upper-tail dependence; the 
survival-Clayton is a rotation of the Clayton by $180^\circ$ and has 
positive upper-tail ($0.87$ in Fig. \ref{fig_cop_pedag1}D)
and $0$ lower-tail dependence. For our model selection procedure (see below), we used 
several families of copulas exhibiting various 
lower- and upper-tail dependence characteristics 
(Clayton, survival-Clayton, Gumbel, survival-Gumbel, Joe, survival-Joe, Frank,
BB1, survival-BB1,
BB6, survival-BB6, BB7, survival-BB7, BB8, survival-BB8, henceforth
abbreviated C, SC, G, SG, J, SJ, F, BB1, SBB1, BB6, SBB6, BB7, SBB7, BB8, SBB8).
Copulas are implemented in the R packages `copula` and `VineCopula`. 
Visualizations are in 
Figs \ref{SM-figped_N}-\ref{SM-figped_SBB8}. 
If LT and UT denote 
lower- and upper-tail dependence measures, their difference measures asymmetry of 
tail dependence.
See @nelsen2006_copula, @joe2014_dependence for further background.

We tested the observational hypothesis (Fig. \ref{MasterFigure}, O) 
principally via a model selection procedure in which the 
copulas listed above were fitted to 
our ecological datasets and fits were compared via the Akaike and Bayesian 
Information Criteria (AIC and BIC). For bivariate datasets 
$(x_i,y_i)$ for $i=1,...,n$, model selection involved several steps. 
First, we produced normalized ranks 
$u_i$ and $v_i$ as in the Introduction. 
Second, we tested the independence of the $u_i$ and $v_i$ 
(using `BiCopIndTest` in the `VineCopula` package). 
Our model 
selection algorithms were ineffective if data could not be distinguished from independent data 
since many of the copula families we considered include the independent copula. 
If independence could be rejected ($0.05$ significance level), model 
selection proceeded. Third, we fit the 16 bivariate copula families listed previously to the 
normalized ranks via maximum likelihood, using `BiCopEst` in `VineCopula`, 
then obtaining AIC and BIC values and accompanying model weights $\text{AIC}_{\text{w}}$ and 
$\text{BIC}_{\text{w}}$ [@burnham2003_modelselection]. `BiCopEst` also provided the lower- and 
upper-tail dependence of the best-fitting member of each family. 
Fourth, $\text{AIC}_{\text{w}}$ values were used to get
model-averaged lower- and upper-tail dependence values using standard model averaging 
formulas [@burnham2003_modelselection]; likewise for BIC.
Model selection methods give the *relative* support of several models, but do not 
indicate whether any of the models are an objectively good fit. To test that, 
and as our final step, we tested the goodness of fit of our AIC-best copula family using a 
bootstrapping procedure (`BiCopGofTest` of `VineCopula`). The procedure performed 
one test based on a Cramer-von Mises statistic and another based on a 
Kolmogorov-Smirnov statistic. To keep computation times reasonable, 
an initial run using $100$ bootstraps was performed; if
the $p$-value from either test was less than $0.2$, tests were re-run with 
$1000$ boostraps.

The first part of the observational hypothesis was tested for a bivariate dataset 
by comparing the AIC weights of the non-N copula
families with that of the N family. 
The second part was tested by comparing 
the model-averaged lower- and upper-tail dependence
values to the lower- and upper-tail dependence values of an N copula (i.e., 0). 
The third part 
was tested by comparing the difference of the model-averaged lower- and upper-tail 
dependence values to the analogous difference for an N copula, again 0.
For multivariate datasets, we carried out the bivariate analysis
for all possible pairwise combinations of locations, and 
compared AIC weights and tail-dependence of 
minimum-AIC copulas with that of an N copula. We also counted the number of location pairs 
for which model-averaged lower-tail dependence was 
greater than model-averaged upper-tail dependence 
or vice-versa, and reported the percentage of non-independent 
location pairs showing stronger lower-tail dependence.

@@

To test for asymmetry of dependence in upper and lower portions of distributions,
we used differences $\cor_l-\cor_u$, $\Ps_l-\Ps_u$ and $\Dsq_u-\Dsq_l$
for smaller datasets, (note the opposite order
in the last of these); and $\cor_{0,u_b}-\cor_{1-u_b,1}$, 
$\Ps_{0,u_b}-\Ps_{1-u_b,1}$ and
$\Dsq_{1-u_b,1}-\Dsq_{0,u_b}$ with $u_b$ close 
to $0$ for large datasets.
We tested our statistics in Appendix \ref{SM-Test_npa_stat} (see also Figs \ref{SM-fig_stat_testing35}-\ref{SM-fig_stat_testing1000}). Code and data for this
project is archived at \url{www.github.com/sghosh89/BIVAN}.

@@

Across essentially all our datasets, copula structures were typically non-N and 
commonly showed asymmetric tail 
dependence: the observational hypothesis was supported (Fig. \ref{MasterFigure}). To make results easier to absorb, we first 
present results for an example bivariate dataset, then we summarize results
for all bivariate datasets, then we present results for an example multivariate dataset, then we summarize for all multivariate datasets.

@@

Having demonstrated that non-N copula structure and asymmetric tail dependence are 
common, we addressed the causal hypotheses (Fig. \ref{MasterFigure}, A-D)
using models. We did not systematically 
test the first two parts of A 
because nonlinear and thresold effects of 
environmental variables on ecological variables are already well documented 
(however, see Discussion material about the green spruce aphid). To test 
the third part of hypothesis A, we formulated a model in which 
an environmental variable influences an ecological variable
more strongly in one of its tails, and explored whether the ecological variable
could then exhibit asymmetric tail dependence across space.
Let $E_i(t)$ be an ecological variable 
measured at location $i$ ($i=1,2$) at time $t$.
Assume the dynamics 
$E_i(t+1)=bE_i(t)+g(\epsilon_i(t))+a \delta_i(t)$, where 
the $\delta_i(t)$ are standard-normally distributed 
and independent across time and locations, $a=0.2$,
$b=0.1, -0.1, 0.5, -0.5$ in different simulations, and 
the $(\epsilon_1(t),\epsilon_2(t))$ were drawn, independently through
time, from a bivariate normal distribution with $\var(\epsilon_i)=1$ and 
$\cov(\epsilon_1,\epsilon_2)=0.8$.
The function $g$ describes how the environment $\epsilon_i(t)$ influences
$E_i(t)$. We used $g$ equal to $g_1$ or $g_2$ in different simulations:
$g_1(\epsilon)$ equals $\epsilon$ if $\epsilon<0$ and $0$ otherwise;
$g_2(\epsilon)$ equals $\epsilon$ if $\epsilon>0$ and $0$ otherwise. Thus
$g_1$ represents environmental effects that occur only for negative 
values and $g_2$ only for positive values of $\epsilon$.
Negative values of $b$ correspond to overcompensating dynamics 
and positive values to undercompensating dynamics; larger $|b|$
means slower return to equilibrium after a disturbance. 
Because $(\epsilon_1,\epsilon_2)$
and $(\delta_1,\delta_2)$ have N copula structure, the Moran 
mechanism analyzed below (B on Fig. \ref{MasterFigure}) 
does not operate here.

@@

For each of the three model setups above, for each of several copula families, for each $\tau=0.1,0.2,\ldots,0.9$, and for each of
$50$ replicate simulations, we generated 5000 noise pairs 
$(\epsilon_1(t),\epsilon_2(t))$ from the
bivariate random variable with the given copula family and the given $\tau$.
We then used this noise to drive the model, and retained both the
noise and population values for the final 500 
time steps. For each simulation, the following
statistics were computed for both noises and populations: 
Pearson, Spearman and Kendall correlations, $\cor_l$, $\cor_u$, $\Ps_l$, $\Ps_u$, 
$\Dsq_l$, $\Dsq_u$, $\cor_l - \cor_u$, $\Ps_l - \Ps_u$, and $\Dsq_u - \Dsq_l$. 
Values were plotted against $\tau$ for noises and populations. If causal hypothesis B is
correct, plots should be similar for populations and noises.

Hypothesis C was not tested. Our next 
causal hypothesis, D on Fig. \ref{MasterFigure}, 
is evolutionary, and pertains to bivariate 
trait data across species, e.g., our bird and mammal 
data. This mechanism is a hypothetical explanation for the right tail dependence
observed in those data (Fig. \ref{fig_biv_multi_raw_cop}F,G). 
The hypothesis is that tail dependence occurs in evolutionary changes
in bivariate characters, and gives rise to tail dependence in observed character values
across extant species.
We simulated bivariate character evolution on an estimate of 
the phylogeny, taken from @GenoudIM2018, of $817$ mammal species.
The root character state and the change 
across each branch were randomly chosen from matrices of one million 
bivariate draws from distributions showing one of five distinct types of copula 
structure: 1) extreme or 2) moderate left-tail dependence,
3) symmetric tail dependence, or 4) moderate or 5) extreme
right-tail dependence (Appendix \ref{SM-evol_mech}). 
All distributions had standard-normal marginals and Spearman 
correlation $0.875$ between components, so our simulations assess 
the impact of copula structure only. For each of the five copulas, mammalian character
evolution was simulated 100 times. For each simulation, symmetry of tail dependence 
of the two characters across phylogeny tips was assessed.
We hypothesized that cases 1 and 2 above would yield stronger left- than right-tail 
dependence in tip characters, and cases 4 and 5 would yield the reverse. 
The simulator was written in Python and used version 4.4 of the DendroPy 
package [@DendroPy].

@@

Causal hypothesis B (Fig. \ref{MasterFigure}) was largely confirmed; 
however, similarities between environmental- and 
ecological-variable copula structure were reduced when dynamics were strongly nonlinear.
For simulations using AR(1) models, copula 
statistics were always very similar for noise and populations 
(Fig \ref{fig_Cause4copula_scatter_C_kend_ar1}, 
\ref{SM-fig_Cause4copula_C_kend_ar1}-\ref{SM-fig_Cause4copula_SJ_kend_ar1}),
though there were significant but minor differences for some statistics
and simulations.

@@

For weak noise and our nonlinear model, copula statistics
were again very similar for noise and model outputs 
(Figs \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_0.1}-
\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}).
Though there were again significant but small differences for some statistics
and simulations, the hypothesis was supported that copula structure in the noise translates
to similar but not identical copula structure in output populations. 
Since many ecological models are nonlinear, this result provides the 
expectation that there will be a Moran-effect-like correspondence 
between noise and model-output copula structure for typical ecological dynamics,
as long as noise is small enough that dynamics stay relatively close
to the model equilibrium. Results that hold for "weak noise" in this sense
are common in ecology. 
We repeated the analysis with $r=1.3$, with similar results. The deterministic 
one-habitat-patch Ricker model exhibits a monotonic approach to a stable equilibrium
at $K$ when $r<1$ (undercompensating dynamics, e.g., the value $r=0.5$ used initially), 
but exhibits an oscillatory approach to $K$ when $r>1$ (overcompensating
dynamics, e.g., $r=1.3$).

For strong noise and using our nonlinear model, 
copula statistics, generally speaking, were approximately similar 
between noise and model
outputs; however, similarity was reduced relative to previous results, 
and for a few simulations, tail dependence asymmetry statistics
even had opposite signs for noise and model outputs 
(Figs \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_0.1}-\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}). For instance, using a C copula with a large Kendall $\tau$,
$\cor_l - \cor_u$ was slightly positive for noise, but slightly negative for 
population model outputs (Fig. \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_1}); 
and likewise for SG and SJ copulas (Figs \ref{SM-fig_Cause4copula_SG_kend_sRicker_nsd_1}, 
\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}). Overall,
similarities between noise and model-output copula structure
were more noticable than discrepancies, in spite of nonlinearities.
We repeated the analysis with $r=1.3$. Discrepancies between noise and population
copula structure in this case were often glaring. Apparently noise of standard deviation 1
interacted very strongly with model nonlinearities when the model was in its overcompensatory 
regime.

>>>>

\noindent \emph{The number of words in the main text (excluding abstract, acknowledgements, references, table and figure legends):} 7497

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=3.5in]{./Results/PedagogFig1.pdf}
\caption{(A-D) Bivariate datasets showing diverse relationships between 
variables; many of these datasets nevertheless have the same Pearson (P)
or Spearman (S) correlation coefficients. (C, D) Contours 
of probability density functions (pdf) of Clayton (C) and survival-Clayton 
(D) copulas with parameter 5 (see section on methods for the observational hypothesis), together
with samples from the corresponding distributions. (E,F) Reflections
of C, D, respectively, about a vertical axis. (G) Contours of the
pdf of a normal copula (methods for the observational hypothesis) with 
parameter chosen to
have the same Spearman correlation as A-D, together with a sample.
(H) The portion of the Spearman correlation within a band 
can be computed
for any band (methods for the observational hypothesis) to describe how dependence
varies from one part of the two distributions to another. Diagonal lines
show two bands, the data in the left/lower band showing stronger
dependence than the right/upper band.
Subscripts h and v 
stand for "horizontal" and "vertical". \label{fig_cop_pedag1}}
\end{center}
\end{figure}

Well-developed approaches do exist, however, and are 
applied widely in other fields, 
that provide a description of the dependence 
between two or more variables that can be mathematically proven 
to be complete [@joe2014_dependence; @nelsen2006_copula]. 
The approaches are based on 
the concept of the *copula*, which we here 
introduce. Copula approaches separate the information of a bivariate 
random variable into two distinct parts: the information in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
dependence). From a data perspective, given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), the dependence
structure is revealed by the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The rank
of the smallest element in a set is $1$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$. See, 
for instance,
Fig. \ref{fig_cop_pedag1}C-D (the points), which show the copula structures of 
A-B, respectively. Ranking makes the marginal distributions of 
the underlying random variables uniform, isolating the 
dependence structure.
See @nelsen2006_copula and @joe2014_dependence
for general background. A theorem of @sklar1959_theorem
states that copulas give a unique and complete 
representation of the dependence
between variables.

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=7in]{./MasterFigure.pdf}
\caption{Guide for hypotheses addressed in the text and how each instantiates the
overall hypothesis. Labels are defined 
in the text to refer to parts of the figure. "Causal hypotheses" (i.e., hypotheses about 
causes of copula structure) are labeled with letters A-D. "Consequential 
hypotheses" (i.e., hypotheses about 
consequences of copula structure) are labeled with X-Z. The "observational 
hypothesis" (i.e., our hypothesis about 
copula structure of data) is labeled O.\label{MasterFigure}}
\end{center}
\end{figure}

One way we may expect a copula approach 
to better reflect ecological relationships relates to Liebig's law of the
minimum. If, for instance, the growth of a plant depends on soil 
nitrogen, N, and other factors, a plot of plant growth rates versus N may 
look more like Fig. \ref{fig_cop_pedag1}A or C than like 
Fig. \ref{fig_cop_pedag1}B or D. Nitrogen 
will control plant growth, producing a clear
relationship, only when it is limiting, in the left portions of the distributions. 
The portion of the distributions that dominate the association is visible in the copula
approach, but not via standard techniques. We therefore frame "causal hypothesis" A 
(labeled in Fig. \ref{MasterFigure}): 
copulas of dependencies between environmental and 
ecological variables will sensitively reveal 
nonlinear, asymmetric and threshold effects for population, community, 
and ecosystem functioning variables; copulas between 
environmental and ecological variables
will commonly show tail dependence; such tail dependence will lead to 
tail dependence across space in ecological variables.

@@

Our causal and consequential hypotheses imply, or rely on, an "observational
hypothesis" (O in Fig. \ref{MasterFigure}), which we here frame. Datasets in 
ecology will commonly have copula structure
distinct from that of a Gaussian/normal distribution (here called 
non-normal copula structure); they will commonly show tail dependence 
distinct from that of a normal distribution; in particular they will
show asymmetric tail dependence.

@@

Conceptually, a bivariate copula is a random variable 
on the unit square, $[0,1] \times [0,1]$, with uniform maginals. 
If $(X,Y)$ is a continuous random
variable and the cummulative distribution functions of $X$ and $Y$ are 
$F_X(x)=P[X\leq x]$ and $F_Y(y)=P[Y\leq y]$, then the dependence structure
of $(X,Y)$ is revealed by the random variable $(F_X(X),F_Y(Y))$,
which takes values in $[0,1] \times [0,1]$ and has 
uniform marginals.

We introduce the copula families we used, and 
measures of tail dependence.
The normal copula family (henceforth N) is a one-parameter
family, with parameter corresponding to the correlation of the related 
bivariate normal distribtion, and controlling the degree of dependence 
(Fig. \ref{fig_cop_pedag1}G for an example).
Left- and right-tail dependence are $0$ for the N copula, and this 
serves as a benchmark for tail dependence in other copulas - positive values of our
tail dependence measures correspond to dependence stronger than an N
copula. The Clayton and 
survival-Clayton families (Fig. \ref{fig_cop_pedag1}C-D) are two more one-parameter 
families. The Clayton has positive lower-tail dependence ($0.87$ for
the example of Fig. \ref{fig_cop_pedag1}C) and $0$ upper-tail dependence; the 
survival-Clayton is a rotation of the Clayton by $180^\circ$ and has 
positive upper-tail ($0.87$ in Fig. \ref{fig_cop_pedag1}D)
and $0$ lower-tail dependence. For our model selection procedure (see below), we used 
several families of copulas exhibiting various 
lower- and upper-tail dependence characteristics 
(Clayton, survival-Clayton, Gumbel, survival-Gumbel, Joe, survival-Joe, Frank,
BB1, survival-BB1,
BB6, survival-BB6, BB7, survival-BB7, BB8, survival-BB8, henceforth
abbreviated C, SC, G, SG, J, SJ, F, BB1, SBB1, BB6, SBB6, BB7, SBB7, BB8, SBB8).
Copulas are implemented in the R packages `copula` and `VineCopula`. 
Visualizations are in 
Figs \ref{SM-figped_N}-\ref{SM-figped_SBB8}. 
If LT and UT denote 
lower- and upper-tail dependence measures, their difference measures asymmetry of 
tail dependence.
See @nelsen2006_copula, @joe2014_dependence for further background.

We tested the observational hypothesis (Fig. \ref{MasterFigure}, O) 
principally via a model selection procedure in which the 
copulas listed above were fitted to 
our ecological datasets and fits were compared via the Akaike and Bayesian 
Information Criteria (AIC and BIC). For bivariate datasets 
$(x_i,y_i)$ for $i=1,...,n$, model selection involved several steps. 
First, we produced normalized ranks 
$u_i$ and $v_i$ as in the Introduction. 
Second, we tested the independence of the $u_i$ and $v_i$ 
(using `BiCopIndTest` in the `VineCopula` package). 
Our model 
selection algorithms were ineffective if data could not be distinguished from independent data 
since many of the copula families we considered include the independent copula. 
If independence could be rejected ($0.05$ significance level), model 
selection proceeded. Third, we fit the 16 bivariate copula families listed previously to the 
normalized ranks via maximum likelihood, using `BiCopEst` in `VineCopula`, 
then obtaining AIC and BIC values and accompanying model weights $\text{AIC}_{\text{w}}$ and 
$\text{BIC}_{\text{w}}$ [@burnham2003_modelselection]. `BiCopEst` also provided the lower- and 
upper-tail dependence of the best-fitting member of each family. 
Fourth, $\text{AIC}_{\text{w}}$ values were used to get
model-averaged lower- and upper-tail dependence values using standard model averaging 
formulas [@burnham2003_modelselection]; likewise for BIC.
Model selection methods give the *relative* support of several models, but do not 
indicate whether any of the models are an objectively good fit. To test that, 
and as our final step, we tested the goodness of fit of our AIC-best copula family using a 
bootstrapping procedure (`BiCopGofTest` of `VineCopula`). The procedure performed 
one test based on a Cramer-von Mises statistic and another based on a 
Kolmogorov-Smirnov statistic. To keep computation times reasonable, 
a run using $100$ bootstraps was performed; if
the $p$-value from either test was less than $0.2$, tests were re-run with 
$1000$ boostraps.

The first part of the observational hypothesis was tested for a bivariate dataset 
by comparing the AIC weights of the non-N copula
families with that of the N family. 
The second part was tested by comparing 
the model-averaged lower- and upper-tail dependence
values to the lower- and upper-tail dependence values of an N copula (i.e., 0). 
The third part 
was tested by comparing the difference of the model-averaged lower- and upper-tail 
dependence values to the analogous difference for an N copula, again 0.
For multivariate datasets, we performed the bivariate analysis
for all possible pairwise combinations of locations, and 
compared AIC weights and tail-dependence of 
minimum-AIC copulas with that of an N copula. We also counted the number of location pairs 
for which model-averaged lower-tail dependence was 
greater than model-averaged upper-tail dependence 
or vice-versa, and reported the percentage of non-independent 
location pairs showing stronger lower-tail dependence.

@@

To test for asymmetry of dependence in upper and lower portions of distributions,
we used differences $\cor_l-\cor_u$, $\Ps_l-\Ps_u$ and $\Dsq_u-\Dsq_l$
for smaller datasets (note the opposite order
in the last of these); and $\cor_{0,u_b}-\cor_{1-u_b,1}$, 
$\Ps_{0,u_b}-\Ps_{1-u_b,1}$ and
$\Dsq_{1-u_b,1}-\Dsq_{0,u_b}$ with $u_b$ close 
to $0$ for large datasets.
We tested our statistics in Appendix \ref{SM-Test_npa_stat} (see also Figs \ref{SM-fig_stat_testing35}-\ref{SM-fig_stat_testing1000}). Code and data for this
project is archived at \url{www.github.com/sghosh89/BIVAN}.

@@

Across essentially all our datasets, copula structures were typically non-N and 
commonly showed asymmetric tail 
dependence: the observational hypothesis (Fig. \ref{MasterFigure}) was supported. To make results easier to absorb, we first 
present results for an example bivariate dataset, then we summarize results
for all bivariate datasets, then we present results for an example multivariate dataset, then we summarize for all multivariate datasets.

@@

Having demonstrated that non-N copula structure and asymmetric tail dependence are 
common, we addressed the causal hypotheses (Fig. \ref{MasterFigure}, A-D)
using models. We did not systematically 
test the first two parts of A 
because nonlinear and thresold effects of 
environmental variables on ecological variables are already well documented 
(however, see Discussion material about the green spruce aphid). To test 
the third part of hypothesis A, we formulated a model in which 
an environmental variable influences an ecological variable
more strongly in one of its tails, and explored whether the ecological variable
could then exhibit asymmetric tail dependence across space.
Let $E_i(t)$ be an ecological variable 
measured at location $i$ ($i=1,2$) at time $t$.
Assume the dynamics 
$E_i(t+1)=bE_i(t)+g(\epsilon_i(t))+a \delta_i(t)$, where 
the $\delta_i(t)$ are standard-normally distributed 
and independent across time and locations, $a=0.2$,
$b=0.1, -0.1, 0.5, -0.5$ in different simulations, and 
the $(\epsilon_1(t),\epsilon_2(t))$ were drawn, independently through
time, from a bivariate normal distribution with $\var(\epsilon_i)=1$ and 
$\cov(\epsilon_1,\epsilon_2)=0.8$.
The function $g$ describes how the environment $\epsilon_i(t)$ influences
$E_i(t)$. We used $g$ equal to $g_1$ or $g_2$ in different simulations:
$g_1(\epsilon)$ equals $\epsilon$ if $\epsilon<0$ and $0$ otherwise;
$g_2(\epsilon)$ equals $\epsilon$ if $\epsilon>0$ and $0$ otherwise. Thus
$g_1$ represents environmental effects that occur only for negative 
values and $g_2$ only for positive values of $\epsilon$.
Negative values of $b$ correspond to overcompensating dynamics 
and positive values to undercompensating dynamics; larger $|b|$
means slower return to equilibrium after a disturbance. 
Because $(\epsilon_1,\epsilon_2)$
and $(\delta_1,\delta_2)$ have N copula structure, the Moran 
mechanism analyzed below (Fig. \ref{MasterFigure}, B) 
does not operate here.

@@

For each of the model setups above, for each of several copula families, for each $\tau=0.1,0.2,\ldots,0.9$, and for each of
$50$ replicate simulations, we generated 5000 noise pairs 
$(\epsilon_1(t),\epsilon_2(t))$ from the
bivariate random variable with the given copula family and the given $\tau$.
We then used this noise to drive the model, and retained both the
noise and population values for the final 500 
time steps. For each simulation, the following
statistics were computed for both noises and populations: 
Pearson, Spearman and Kendall correlations, $\cor_l$, $\cor_u$, $\Ps_l$, $\Ps_u$, 
$\Dsq_l$, $\Dsq_u$, $\cor_l - \cor_u$, $\Ps_l - \Ps_u$, and $\Dsq_u - \Dsq_l$. 
Values were plotted against $\tau$ for noises and populations. If causal hypothesis B is
correct, plots should be similar for populations and noises.

Hypothesis C was not tested. Our next 
causal hypothesis, Fig. \ref{MasterFigure}, D, 
is evolutionary, and pertains to bivariate 
trait data across species, e.g., our bird and mammal 
data. This mechanism is a hypothetical explanation for the right tail dependence
observed in those data (Fig. \ref{fig_biv_multi_raw_cop}F,G). 
The hypothesis is that tail dependence occurs in evolutionary changes
in bivariate characters, and gives rise to tail dependence in observed character values
across extant species.
We simulated bivariate character evolution on an estimate of 
the phylogeny, taken from @GenoudIM2018, of $817$ mammal species.
The root character state and the change 
across each branch were randomly chosen from matrices of one million 
bivariate draws from distributions showing one of five distinct types of copula 
structure: 1) extreme or 2) moderate left-tail dependence,
3) symmetric tail dependence, or 4) moderate or 5) extreme
right-tail dependence (Appendix \ref{SM-evol_mech}). 
All distributions had standard-normal marginals and Spearman 
correlation $0.875$ between components, so our simulations assess 
the impact of copula structure only. For each of the five copulas, mammalian character
evolution was simulated 100 times. For each simulation, symmetry of tail dependence 
of the two characters across phylogeny tips was assessed.
We hypothesized that cases 1 and 2 above would yield stronger left- than right-tail 
dependence in tip characters, and cases 4 and 5 would yield the reverse. 
The simulator was written in Python and used version 4.4 of the DendroPy 
package [@DendroPy].

@@

Causal hypothesis B (Fig. \ref{MasterFigure}) was largely confirmed; 
however, similarities between environmental- and 
ecological-variable copula structure were reduced when dynamics were strongly nonlinear.
For simulations using AR(1) models, copula 
statistics were always very similar for noise and populations 
(Fig \ref{fig_Cause4copula_scatter_C_kend_ar1}, 
\ref{SM-fig_Cause4copula_C_kend_ar1}-\ref{SM-fig_Cause4copula_SJ_kend_ar1}),
though there were significant but minor differences for some statistics
and simulations.
For weak noise and our nonlinear model, copula statistics
were again very similar for noise and model outputs 
(Figs \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_0.1}-
\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}).
Though there were again significant but small differences for some statistics
and simulations, the hypothesis was supported that copula structure in the noise translates
to similar but not identical copula structure in output populations. 
Since many ecological models are nonlinear, this result provides the 
expectation that there will be a Moran-effect-like correspondence 
between noise and model-output copula structure for typical ecological dynamics,
as long as noise is small enough that dynamics stay relatively close
to the model equilibrium. Results that hold for "weak noise" in this sense
are common in ecology. 
We repeated the analysis with $r=1.3$, with similar results. The deterministic 
one-habitat-patch Ricker model exhibits a monotonic approach to a stable equilibrium
at $K$ when $r<1$ (undercompensating dynamics, e.g., the value $r=0.5$ used initially), 
but exhibits an oscillatory approach to $K$ when $r>1$ (overcompensating
dynamics, e.g., $r=1.3$).

@@

For strong noise and using our nonlinear model, 
copula statistics, generally speaking, were approximately similar 
between noise and model
outputs; however, similarity was reduced relative to previous results, 
and for a few simulations, tail dependence asymmetry statistics
even had opposite signs for noise and model outputs 
(Figs \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_0.1}-\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}). For instance, using a C copula with a large Kendall $\tau$,
$\cor_l - \cor_u$ was slightly positive for noise, but slightly negative for 
population model outputs (Fig. \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_1}); 
and likewise for SG and SJ copulas (Figs \ref{SM-fig_Cause4copula_SG_kend_sRicker_nsd_1}, 
\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}). Overall,
similarities between noise and model-output copula structure
were more noticable than discrepancies, in spite of nonlinearities.
We repeated the analysis with $r=1.3$. Discrepancies between noise and population
copula structure in this case were often glaring. Apparently noise of standard deviation 1
interacted strongly with model nonlinearities when the model was in its overcompensatory 
regime.

For our model, hypothesis D (Fig. \ref{MasterFigure}) was 
correct that asymmetric tail dependence in evolutionary changes 
can produce similarly asymmetric tail dependence in character distributions across
phylogeny tips (Appendix \ref{SM-evol_mech}, Fig. \ref{SM-fig_cop_evo} for details). This topic 
is revisited in the Discussion.
